Stool Microbiome Analysis

### 2022-03-24 Rmarkdown/analysis/Fecal_Microbiome_TMAO_Analysis_211105.Rmd
Limit to TMAO analyses

1) Plots per section [expand as analysis goes final]
2) Tables [expand as analysis goes final]

@@

Table of read counts at the Phylum level. We will filter low abundant Phyla. We will also require that taxa are present in at least 10% of samples (at least 36 samples). We will then agglomerate at the Genus level.

@@

## Family Plot of TMAO Quantile

```{r Phylo Abundance Plot by TMAO Quantile}
# Agglomerate taxa
PSOtmao1F = tax_glom(PSOtmao1, "Family", NArm = TRUE)
PSOtmao1F.top10 <- microbiome::aggregate_top_taxa(PSOtmao1F, "Family", top = 10)

ps1F.10.family.comp <- microbiome::transform(PSOtmao1F.top10, transform="compositional")

plot.composition.relAbun <- microbiome::plot_composition(ps1F.10.family.comp,
                                             sample.sort = "Description",
                                             x.label = "tmao_quantile",
                                             group_by = "tmao_quantile",
                                             average_by = "tmao_quantile") +
  theme_minimal() +
  scale_fill_brewer("Family", palette = "Paired") + 
  ggtitle("Relative Abundance at Family Level")

plot.composition.relAbun
```

@@

From Dr. Kable's example,

> DESeq2 Set Up
> Test the impact of fiber cluster, accounting for covariates, using the log ratio test
>
> dds.ratio.batcho = phyloseq_to_deseq2(ps.genus, ~ Batch + k4_cluster)
> This performs a likelihood ratio test, which determines if the increased likelihood of the data in the full model, where we include all terms, is more than expected if the terms excluded in the reduced model are really zero. 
>
> dds.ratio.batcho <- DESeq(dds.ratio.batcho, test="LRT", reduced = ~ Batch )
> 
> The p-value is based on the LRT (so the same for everything), but the contrasts give the right log2 fold change.
> 
> resultsNames(dds.ratio.batcho)
> res.1v2.batcho <- results(dds.ratio.batcho, contrast=c("k4_cluster", "2", "1"))

@@

## Factored by TMAO Quantile

```{r DESeq2 analysis - TMAO quantile}
# DESeq analysis - tmao_quantile
mb <- phyloseq_to_deseq2(PSOtmao1G, ~ sex + age + bmi_final.x + tmao_quantile)
mb.ratio <- DESeq(mb, test = "LRT", reduced = ~ sex + age + bmi_final.x) # remove variable of interest here, TMAO
res.ratio <- results(mb.ratio, contrast = c("tmao_quantile", "Quantile1", "Quantile4"))
alpha = 0.05
sigtab = res.ratio[which(res.ratio$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(PSOtmao1G)[rownames(sigtab), ], "matrix"))
sigtab$family_genus <- paste0(sigtab$Family, sigtab$Genus)
head(sigtab)

#~~~~~~~~~~~~~~~
# Plot Results
#~~~~~~~~~~~~~~~
resOTU <- rownames(sigtab)
otuData <- as.data.frame(otu_table(PSOtmao1G))
otuData_desRes <- otuData[rownames(otuData) %in% resOTU,]
#otuData_desRes <- as.data.frame(t(otuData_desRes))
# Get sample data groupings
sampleData <- as.data.frame(sample_data(PSOtmao1G))
sampleData <- subset(sampleData, select = c(tmao_mdn, tmao_quantile, tmao_tertile, tmao))
# Get tax table
taxaData <- as.data.frame(tax_table(PSOtmao1G))
taxaData_res <- taxaData[rownames(taxaData) %in% resOTU,]
taxaData_res$OrderFamilyGenus <- paste0(taxaData_res$Order, taxaData_res$Family, taxaData_res$Genus)

# Merge
ggData <- merge(taxaData_res, otuData_desRes, by = 0)
rownames(ggData) <-  ggData$OrderFamilyGenus
ggData <- ggData[,-(1:9)]
ggData <- as.data.frame(t(ggData))
ggData <- merge(ggData, sampleData, by = 0)
rownames(ggData) <- ggData[,"Row.names"]
ggData <- subset(ggData, select = -c(Row.names))

# Loop Plot
nres <- length(resOTU)
for (i in 1:nres) { 
    print(ggplot(ggData, aes(x = tmao_quantile, y = log(ggData[,i]))) +
            geom_boxplot()+
            ylab(colnames(ggData)[i])) 
}

#~~~~~~~~~~~~~~~
# Plot Results Fold Change
#~~~~~~~~~~~~~~~
sigtabgen = subset(sigtab, !is.na(Genus))
# Phylum order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Phylum, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Phylum = factor(as.character(sigtabgen$Phylum), levels=names(x))
# Genus order
x = tapply(sigtabgen$log2FoldChange, sigtabgen$Genus, function(x) max(x))
x = sort(x, TRUE)
sigtabgen$Genus = factor(as.character(sigtabgen$Genus), levels=names(x))
ggplot(sigtabgen, aes(y=family_genus, x=log2FoldChange, color=Phylum)) + 
  geom_vline(xintercept = 0.0, color = "gray", size = 0.5) +
  geom_point(size=3) + 
  ylab("Taxonomy") +
  xlab("Log2 Fold Change") + 
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5))
```

@@

Play around with the coloring to try to figure out what is discriminating these groups!

```{r What defines the PCoA groups exploration}
colnames(sample_data(PSOtmao1G))
loopPheno <- c("pa_ee_total_mdn","endo_ln_rhi_mdn", "endo_al75_mdn", "betaine_mdn", "carnitine_mdn", "tmao_mdn",  "choline_mdn", "creatinine_mdn", "phosphocholine_mdn", "hei_asa24_totalveg_mdn", "hei_asa24_totalproteinfoods_mdn", "hei_asa24_sodium_mdn", "hei_asa24_refinedgrains_mdn", "hei_asa24_sfat_mdn",  "hei_asa24_addsug_mdn", "hei_asa24_wholegrain_mdn", "hei_asa24_totalscore_mdn", "tmao_log_mdn", "TMAO_5", "TMAO_6", "TMAO_7", "sex", "age_cat", "bmi_cat", "tmao_quantile", "tmao_tertile")

# PCoA_WU, PCoA_UnWU, PCoA_BC plug in to view
for(i in 1:length(loopPheno)){
print(plot_ordination(PSOtmao1G, PCoA_WU, 
                    color = loopPheno[i],
                    shape = NULL,
                    axes = 1:2) +
  #scale_color_brewer(palette = "Dark2", labels = c("Below", "Above")) +
  stat_ellipse(linetype = 1) +
  labs(color = loopPheno[i]) +
  theme_minimal() + 
  theme(axis.text = element_text(size = 12), # x and y axis text size
        axis.title = element_text(size = 12), # x and y label text size
        legend.text = element_text(size = 12), # legend text size
        legend.title = element_text(size = 12), # label legend text size
        axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5)))
}

# From the plots, Weighted UniFrac and endo_al75 and betaine look different
bdis <- betadisper(dismax_wu, df$betaine_mdn)
permutest(bdis) # NS, good
boxplot(bdis, las = 2)

# ADONIS (perMANOVA) analysis
adonis2(dismax_wu ~ betaine + sex + age + bmi_final.y, data=df, method = "wunifrac", perm=999) # different

# ANCOM

## ANCOM2 no-covariate analysis

```{r ANCOM - genus level}
source("../../Rscripts/ancom_v2.1.R")

# OTU table
MicroData <- otu_table(PSOtmao1G)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao1G)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
MetaData2$Sample.ID <- rownames(MetaData2)

feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
# Function
prepro = feature_table_pre_process(feature_table = feature_table,
                                   meta_data = MetaData2,
                                   sample_var = sample_var,
                                   group_var = NULL,
                                   out_cut = out_cut,
                                   zero_cut = zero_cut,
                                   lib_cut = lib_cut,
                                   neg_lb = neg_lb)

# Step 2: ANCOM
feature_table = prepro$feature_table 
meta_data = prepro$meta_data
struc_zero = prepro$structure_zeros

# Define variables for ANCOM
main_var = "tmao_mdn"; p_adj_method = "BH"; alpha = 0.05
adj_formula = NULL ; rand_formula = NULL
# Run ANCCOM
t_start = Sys.time()
res = ANCOM(feature_table = feature_table, 
            meta_data = meta_data, 
            struc_zero = struc_zero, 
            main_var = main_var, 
            p_adj_method = p_adj_method, 
            alpha = alpha, 
            adj_formula = adj_formula, 
            rand_formula = rand_formula)
t_end = Sys.time()
t_run = t_end - t_start 
t_run # around 5s

# Results
res$out
res$fig$data

# Step 3: Volcano Plot
# Number of taxa except structural zeros
n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")

# Annotation data
dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")

fig = res$fig +  
  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
fig

# ORGANIZE PLOT

TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
rownames(TaxaData2) <- TaxaData2$Feature.ID

ancomRes <- res$out

test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)

# Make plot 
tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
asv_id <- unique(tempdata$taxa_id)

MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]

# Rename to the taxonomy 
#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
# Make a unique naming column that incorporates multiple levels of taxonomy
TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
rownames(TaxaData3) <- rownames(TaxaData2)
# Merge
MicroData5 <- merge(MicroData4, TaxaData3,
                    by=0, all.x = T, all.y = F)

row.names(MicroData5) <- MicroData5$OrderFamilyGenus

MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))

# Merge with the phenotype data 
MetaData3 <- MetaData2 %>% dplyr::select(tmao_mdn)

ggData <- merge(MetaData3, MicroData5, by =0)
row.names(ggData) <- ggData$Row.names
ggData <- ggData %>% dplyr::select(-Row.names)

# Gather the data
library(tidyr)
ggData2 <- gather(ggData,
                  key = "Taxa",
                  value = "Abundance",
                  -tmao_mdn)

ggData2$Abundance <- log(ggData2$Abundance)

# Plot it
p <- ggplot(data = ggData2,
            aes(x = Taxa, 
                y = Abundance,
                fill = tmao_mdn), 
            alpha = 0.1) +
  geom_boxplot(lwd=.25) +
  xlab(NULL) +
  ylab("Log of Taxa Abundance") +
  theme_bw() + # note order matters here - need bw to be before theme()
  theme(axis.text.y = element_text( hjust = 1),
        axis.text.x = element_text(angle = 60, hjust = 1),
        text=element_text(size=11))  +
  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
  #scale_fill_viridis(discrete = TRUE) +
  #coord_flip() +
  labs(fill = "TMAO \n Classification")

p
```

## ANCOM2 covariate analysis

ANCOM2 lets you include covariates into the analysis. If we add the same covariates as in the DESeq2 analysis, do we get similar results?

```{r ANCOM - genus level with covariates median}
source("../../Rscripts/ancom_v2.1.R")

# OTU table
MicroData <- otu_table(PSOtmao1G)
MicroData <- as.data.frame(as.matrix(MicroData), stringsAsFactors = F)
# Add 1
#MicroData2 <- MicroData +1
#MicroData2 <- as.data.frame(as.matrix(t(MicroData2),stringsAsFactors = F))
#dim(MicroData2) # 103 x 98

# Load the phenotype data
#------------------------
# Perform ANCOM analysis
# Step 1: Data preprocessing
# OTU data
otu_data <- otu_table(PSOtmao1G)
otu_data <- as.data.frame(as.matrix(otu_data), stringsAsFactors = F)
otu_id <- rownames(otu_data)

# Metadata 
#df <- readRDS("../data_processed/mapping/Biocrates_mapping_BD1_210809.rds")
#df$p.Cresol.SO4_bd1_mdn <- plyr::revalue(df$p.Cresol.SO4_bd1_mdn, c("0" = "< Median", "1" = "> Median"))
# Subset
MetaData2 <- df2 %>% dplyr::select(c(tmao_mdn, age, sex, bmi_final))
MetaData2$Sample.ID <- rownames(MetaData2)

feature_table = otu_data; sample_var = "Sample.ID"; group_var = NULL
out_cut = 0.05; zero_cut = 0.95; lib_cut = 1000; neg_lb = FALSE # make zero_cut=0.95 to match criteria used for other metabolites
# Function
prepro = feature_table_pre_process(feature_table = feature_table,
                                   meta_data = MetaData2,
                                   sample_var = sample_var,
                                   group_var = NULL,
                                   out_cut = out_cut,
                                   zero_cut = zero_cut,
                                   lib_cut = lib_cut,
                                   neg_lb = neg_lb)

# Step 2: ANCOM
feature_table = prepro$feature_table 
meta_data = prepro$meta_data
struc_zero = prepro$structure_zeros

# Define variables for ANCOM
main_var = "tmao_mdn"; p_adj_method = "BH"; alpha = 0.05
adj_formula = c("age + sex + bmi_final"); rand_formula = NULL
# Run ANCCOM
t_start = Sys.time()
res = ANCOM(feature_table = feature_table, 
            meta_data = meta_data, 
            struc_zero = struc_zero, 
            main_var = main_var, 
            p_adj_method = p_adj_method, 
            alpha = alpha, 
            adj_formula = adj_formula, 
            rand_formula = rand_formula)
t_end = Sys.time()
t_run = t_end - t_start 
t_run # around 5s

# Results
res$out
res$fig$data

# Step 3: Volcano Plot
# Number of taxa except structural zeros
n_taxa = ifelse(is.null(struc_zero), nrow(feature_table), sum(apply(struc_zero, 1, sum) == 0))
# Cutoff values for declaring differentially abundant taxa
cut_off = c(0.9 * (n_taxa -1), 0.8 * (n_taxa -1), 0.7 * (n_taxa -1), 0.6 * (n_taxa -1))
names(cut_off) = c("detected_0.9", "detected_0.8", "detected_0.7", "detected_0.6")

# Annotation data
dat_ann = data.frame(x = min(res$fig$data$x), y = cut_off["detected_0.7"], label = "W[0.7]")

fig = res$fig +  
  geom_hline(yintercept = cut_off["detected_0.7"], linetype = "dashed") + 
  geom_text(data = dat_ann, aes(x = x, y = y, label = label), 
            size = 4, vjust = -0.5, hjust = 0, color = "orange", parse = TRUE)
fig

# ORGANIZE PLOT

TaxaData2 <- readRDS("../../data/processed/microbiome/phyloseq_inputs/merged-taxonomy-dada2-retrained.rds")
rownames(TaxaData2) <- TaxaData2$Feature.ID

ancomRes <- res$out

test <- merge(ancomRes, TaxaData2, by.x = "taxa_id", by.y = "Feature.ID", all.x = TRUE, all.y = FALSE)

# Make plot 
tempdata <- test[test$detected_0.6 == TRUE,] # 0.6 but can make 0.7
asv_id <- unique(tempdata$taxa_id)

MicroData4 <- feature_table[row.names(feature_table) %in% asv_id,]

# Rename to the taxonomy 
#MicroData5 <- as.data.frame(as.matrix(t(MicroData4)))
TaxaData3 <- TaxaData2 %>% dplyr::select(Order, Family, Genus)
# Make a unique naming column that incorporates multiple levels of taxonomy
TaxaData3$OrderFamilyGenus <- paste(TaxaData3$Order, TaxaData3$Family, TaxaData3$Genus)
rownames(TaxaData3) <- rownames(TaxaData2)
# Merge
MicroData5 <- merge(MicroData4, TaxaData3,
                    by=0, all.x = T, all.y = F)

row.names(MicroData5) <- MicroData5$OrderFamilyGenus

MicroData5 <- MicroData5 %>%  dplyr::select(-Order, -Family, -Genus, -OrderFamilyGenus, -Row.names)
MicroData5 <- as.data.frame(as.matrix(t(MicroData5)))

# Merge with the phenotype data 
MetaData3 <- MetaData2 %>% dplyr::select(tmao_mdn)

ggData <- merge(MetaData3, MicroData5, by =0)
row.names(ggData) <- ggData$Row.names
ggData <- ggData %>% dplyr::select(-Row.names)

# Gather the data
library(tidyr)
ggData2 <- gather(ggData,
                  key = "Taxa",
                  value = "Abundance",
                  -tmao_mdn)

ggData2$Abundance <- log(ggData2$Abundance)

# Plot it
p <- ggplot(data = ggData2,
            aes(x = Taxa, 
                y = Abundance,
                fill = tmao_mdn), 
            alpha = 0.1) +
  geom_boxplot(lwd=.25) +
  xlab(NULL) +
  ylab("Log of Taxa Abundance") +
  theme_bw() + # note order matters here - need bw to be before theme()
  theme(axis.text.y = element_text( hjust = 1),
        axis.text.x = element_text(angle = 60, hjust = 1),
        text=element_text(size=11))  +
  scale_fill_manual(values=c("#2D708EFF", "#B8DE29FF")) + 
  #scale_fill_viridis(discrete = TRUE) +
  #coord_flip() +
  labs(fill = "TMAO \n Classification")

p
```
# Stats for paper

@@

How many OTUs are present in 5%, 25%, 50% of samples?

```{r}

@@

THIS CHUNK WORK IN PROGRESS

>>>>

1) Plots of differential abundance, and alpha and beta diversity results.
2) Tables used in the supplemental files of the manuscript.

@@

Table of read counts at the Phylum level. We will filter low abundant Phyla. We will also require that taxa are present in at least 5% of samples. We will then agglomerate at the Genus level.

@@

@@

@@

@@

# Stats for paper

@@

@@
